Hilbert space methods to approximate Gaussian processes using Stan

### 2019-04-10 Paper/manuscript/manuscript_05.tex
two minor fixes
add 2D simulate study case and improve some figures

Given $n$ observations in the data, the computational complexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This would limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. In the unidimensional case the spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{align}\label{eq:specdens}
S_{\nu=\infty}(w)&= \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \exp\left(-0.5 \ell^2 w^2\right), \\
S_{\nu=\frac{1}{2}}(w)&= 2\sigma^2 \frac{1}{\ell}\left(\frac{1}{\ell^2} + \omega^2\right)^{-1}, \\
S_{\nu=\frac{3}{2}}(w)&= 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}\left(\frac{3}{\ell^2} + \omega^2\right)^{-2}, \\
%S_{\nu=\frac{5}{2}}(w)&= \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}\left(\frac{5}{\ell^2} + \omega^2\right)^{-3},  
\end{align}

\noindent where variable $w\in \mathbb{R}$ is in the frequency domain, and $\ell\in \mathbb{R}$. By using the separability property of the spectral densities functions, the $D$-dimensional form of these spectral density functions, where $\mathbf{w}=\{w_1,w_2\}$ and $\mathbf{\ell}=\{\ell_1,\ell_2\}$ are
%
\begin{align}\label{eq:specdens_multi}
S_{\nu=\infty}(\bm{\omega})&= \sqrt{2\pi} \cdot \ell_1 \cdot \exp\left(-0.5 \ell_1^2 \omega_1^2\right) \cdot  \sqrt{2\pi} \cdot \ell_2 \cdot \exp\left(-0.5 \ell_2^2 \omega_2^2\right) \\
S_{\nu=\frac{1}{2}}(\bm{\omega})&= 2 \frac{1}{\ell_1}\left(\frac{1}{\ell_1^2} + \omega_1^2\right)^{-1} \cdot 2 \frac{1}{\ell_2}\left(\frac{1}{\ell_2^2} + \omega_2^2\right)^{-1} \\
S_{\nu=\frac{3}{2}}(\bm{\omega})&= 4 \frac{\sqrt{3}}{\ell_1}^{3}\left(\frac{3}{\ell_1^2} + \omega_1^2\right)^{-2} \cdot 4 \frac{\sqrt{3}}{\ell_2}^{3}\left(\frac{3}{\ell_2^2} + \omega_2^2\right)^{-2},
\end{align}

@@

The approximate Gaussian process method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs (Paul: introducing such an abbreviation perhaps shortens what we are going to write in the next sections). Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}.

We begin by focusing on the case of a unidimensional input space (i.e., on GPs with just a single covariate) such that $\Omega \in \{-L,L\} \subset {\rm I\!R}$, where $L$ is some positive real value to which we also refer as boundary condition. As $\Omega$ describes the interval in which the approximations are valid, $L$ plays a critical role in the accuracy of HSGPs. We will come back to this issue in Section 3.

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in Equation (\ref{eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{eqnarray}\label{bf&lambda}
\begin{split}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right).
\end{split}
\end{eqnarray}

@@

\noindent where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x^i)$:
%
\begin{eqnarray}
\Phi &=&  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{eqnarray}

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number $D$-tuples is given by $m^{\ast} = \prod_{d=1}^{D} m_d$, where $m_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!N}^{m^{\ast} \times D}$ be the matrix of all those $D$-tuples. For example, suppose we have $D=3$ dimensions and use $m_{1}=2$, $m_{2}=2$ and $m_{3}=3$ eigenfunctions and eigenvalues for the first, second and third dimension, respectively. Then, the number of multivariate eigenfunctions and eigenvalues is $m^{\ast} = m_{1} \cdot m_{2} \cdot m_{3} = 12$ and the matrix $\mathbb{S}\in {\rm I\!N}^{12 \times 3}$ is given by
\begin{align}\small
\mathbb{S}=
\left[ {\begin{array}{ccc}
1 & 1 & 1 \nonumber \\
1 & 1 & 2 \\
1 & 1 & 3 \\
1 & 2 & 1 \\
1 & 2 & 2 \\
1 & 2 & 3 \\
2 & 1 & 1 \\
2 & 1 & 2 \\
2 & 1 & 3 \\
2 & 2 & 1 \\
2 & 2 & 2 \\
2 & 2 & 3 
\end{array} } \right]
\end{align}

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\lambda^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$. Thus, for $\mathbf{x}^D \in \Omega$ and $j=1,\ldots,m$, we have 
%
\begin{align} \label{bf&lambda_multi}
\begin{split}
\phi^{\ast}_j(\mathbf{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \\
%
\lambda^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. 
\end{split}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{approxcov_multi}
k(\mathbf{x},\mathbf{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\lambda^{\ast}_j}\right)
\phi^{\ast}_j(\mathbf{x}) \phi^{\ast}_j(\mathbf{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf}
f(\mathbf{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\lambda^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\mathbf{x}) \beta_j, 
\end{equation}

@@

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equation (\ref{bf&lambda}) for the unidimensional case and in Equation (\ref{bf&lambda_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{bf&lambda}) and (\ref{bf&lambda_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.47, trim = 0mm 0mm 0mm 10mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.47, trim = 25mm 0mm 0mm 10mm, clip]{fig1_Cov_J.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different number of basis functions $m$. (right) Covariance functions of both regular GP model and HSGP model for different number of basis functions $m$.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.47, trim = 0mm 0mm 0mm 10mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.47, trim = 25mm 0mm 0mm 10mm, clip]{fig2_Cov_L.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different boundary factors $c$. (right) Covariance functions of both regular GP model and HSGP model for different boundary factors $c$.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.41, trim = 0mm 3mm 0mm 0mm, clip]{fig4_Post_&_Cov_part1.pdf}}
\subfigure{\includegraphics[scale=0.41, trim = 0mm 3mm 0mm 3mm, clip]{fig4_Post_&_Cov_part2.pdf}}
\subfigure{\includegraphics[scale=0.41, trim = 0mm 0mm 0mm 3mm, clip]{fig4_Post_&_Cov_part3.pdf}}
\caption{Posterior predictive mean functions and covariance functions of both regular GP model and HSGP model for different number of basis functions $m$ and values of the boundary factor $c$.} 
  \label{fig4_Post_&_Cov_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.44]{fig5_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.44]{fig5_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the regular GP model. (left) RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$. (right) RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$. }
  \label{fig5_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.48, trim = 0mm 0mm 5mm 0mm, clip]{fig6_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.48, trim = 31mm 0mm 0mm 0mm, clip]{fig6_magnitud_vs_J.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both regular GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig6_lscale_vs_J}
\end{figure}

@@

\vspace{3mm}
\section{Univariate examples}\label{sec:gp_examples1D}

@@

This example consists of a simulated dataset with 250 draws of a Gaussian process prior with a Mat{\'e}rn covariance function with 3/2 degrees of freedom and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, and inputs values $\{\mathbf{x}_i\}_{i=1}^{250} \in \{-1,1\} \subset {\rm I\!R}$. To form the final dataset $\mathbf{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.

The regular GP model for fitting this simulated dataset $\mathbf{y}$ can be written as follows,
%
\begin{eqnarray*}\label{eq:latentgp_simudata1}
\begin{split}
\mathbf{y} &= f(\mathbf{x}) + \boldsymbol{\epsilon} \\
\boldsymbol{\epsilon} &\sim \mathcal{N}(0, \sigma^2  \mathbf{I}) \\
f(\mathbf{x}) &\sim \mathcal{GP}(0, k(\mathbf{x}, \mathbf{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $\mathbf{I}$ represents the identity matrix. The previous formulation corresponds to the latent form of a GP model, where $f(\mathbf{x})$ represents the underlying functional model to the noisy data, and it is modeled by a GP prior with a Mat{\'e}rn($\nu=3/2$) covariance function $k$ (\ref{covfun}). Saying that the function values $f(\mathbf{x})$ follows a GP model is equivalent to say that $f(\mathbf{x})$ are multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(\mathbf{x}_i,\mathbf{x}_j,\theta)$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using sampling methods such as HMC, would be its marginalized form,
%
\begin{equation*}\label{eq:marginalizedgp_simudata1}
\mathbf{y} \sim \mathcal{N}(0, K + \sigma^2 \mathbf{I} )
\end{equation*}

@@

In the HSGP model, the latent function values $f(x)$ are approximated as (\ref{approxf}),
%
\begin{equation*}
f(x) \approx \sum_{j}^m \left( S(\sqrt{\lambda_j})\right)^{1/2} \phi_j(x) \beta_j,
\end{equation*}

\noindent with the spectral density $S$ (\ref{specdens}) as a function of $\sqrt{\lambda_j}$,
%
\begin{eqnarray*}
S(\sqrt{\lambda_j})&=& 4\alpha^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \lambda_j)^{-2}, 
\end{eqnarray*}

>>>>

Given $n$ observations in the data, the computational complexity and memory requirements of exact GP implementation in general scale as $O(n^3)$ and $O(n^2)$, respectively. This limit their application to rather small data sets of a few tens of thousands observations at most. The problem becomes more severe when performing full Bayesian inference via sampling methods, where in each sampling step we need $O(n^3)$ computations when inverting the Gram matrix of the covariance function, usually through Cholesky factorization. To alleviate these computational demands, several approximate methods have been proposed.

@@

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. In the unidimensional case the spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{align}
S_{\nu=\infty}(w)&= \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \exp\left(-0.5 \ell^2 w^2\right), \label{eq:specdens_inf}  \\
S_{\nu=\frac{1}{2}}(w)&= 2\sigma^2 \frac{1}{\ell}\left(\frac{1}{\ell^2} + \omega^2\right)^{-1}, \label{eq:specdens_12}  \\
S_{\nu=\frac{3}{2}}(w)&= 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}\left(\frac{3}{\ell^2} + \omega^2\right)^{-2}, \label{eq:specdens_32} \\
%S_{\nu=\frac{5}{2}}(w)&= \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}\left(\frac{5}{\ell^2} + \omega^2\right)^{-3}, \label{eq:specdens_52} 
\end{align}

\noindent where variable $w\in \mathbb{R}$ is in the frequency domain, and $\ell\in \mathbb{R}$. By using the separability property of the spectral densities functions, the $D$-dimensional form of these spectral density functions, where $\mathbf{w}=\{w_1,w_2\}$ and $\mathbf{\ell}=\{\ell_1,\ell_2\}$ are
%
\begin{align}
S_{\nu=\infty}(\bm{\omega})&= \sqrt{2\pi} \cdot \ell_1 \cdot \exp\left(-0.5 \ell_1^2 \omega_1^2\right) \cdot  \sqrt{2\pi} \cdot \ell_2 \cdot \exp\left(-0.5 \ell_2^2 \omega_2^2\right) \label{eq:specdens_multi_inf} \\
S_{\nu=\frac{1}{2}}(\bm{\omega})&= 2 \frac{1}{\ell_1}\left(\frac{1}{\ell_1^2} + \omega_1^2\right)^{-1} \cdot 2 \frac{1}{\ell_2}\left(\frac{1}{\ell_2^2} + \omega_2^2\right)^{-1} \label{eq:specdens_multi_12} \\
S_{\nu=\frac{3}{2}}(\bm{\omega})&= 4 \frac{\sqrt{3}}{\ell_1}^{3}\left(\frac{3}{\ell_1^2} + \omega_1^2\right)^{-2} \cdot 4 \frac{\sqrt{3}}{\ell_2}^{3}\left(\frac{3}{\ell_2^2} + \omega_2^2\right)^{-2}, \label{eq:specdens_multi_32}
\end{align}

@@

The approximate Gaussian process method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}.

We begin by focusing on the case of a unidimensional input space (i.e., on GPs with just a single covariate) such that $\Omega \in [-L,L] \subset {\rm I\!R}$, where $L$ is some positive real value to which we also refer as boundary condition. As $\Omega$ describes the interval in which the approximations are valid, $L$ plays a critical role in the accuracy of HSGPs. We will come back to this issue in Section 3.

@@

\noindent The eigenvalues $\lambda_j>0$ are real and positive because the Laplacian is a positive definite Hermitian operator, and the eigenfunctions $\phi_j$ for the eigenvalues problem in Equation (\ref{eigenproblem}) are sinusoidal functions. Independently of the covariance function, they can be computed as
%
\begin{align}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \label{eigenvalue}\\
\phi_j(x)&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(x+L)\right). \label{eigenfunction}
\end{align}

@@

\noindent where $\Phi \in {\rm I\!R}^{n\times m}$ is the matrix of eigenfunctions $\phi_j(x_i)$:
%
\begin{eqnarray}
\Phi &=&  \left[ {\begin{array}{ccc}
   \phi_1(x_1) & \cdots & \phi_m(x_1)  \\
    \vdots &\ddots & \vdots  \nonumber \\ 
    \phi_1(x_n) & \cdots & \phi_m(x_n) \\
  \end{array} } \right].
\end{eqnarray}

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number $D$-tuples is given by 
%
\begin{align} \label{m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

\noindent where $m_d$ is the number of basis function for the dimension $d$. Let $\mathbb{S}\in {\rm I\!N}^{m^{\ast} \times D}$ be the matrix of all those $D$-tuples. For example, suppose we have $D=3$ dimensions and use $m_{1}=2$, $m_{2}=2$ and $m_{3}=3$ eigenfunctions and eigenvalues for the first, second and third dimension, respectively. Then, the number of multivariate eigenfunctions and eigenvalues is $m^{\ast} = m_{1} \cdot m_{2} \cdot m_{3} = 12$ and the matrix $\mathbb{S}\in {\rm I\!N}^{12 \times 3}$ is given by
\begin{align}\small
\mathbb{S}=
\left[ {\begin{array}{ccc}
1 & 1 & 1 \nonumber \\
1 & 1 & 2 \\
1 & 1 & 3 \\
1 & 2 & 1 \\
1 & 2 & 2 \\
1 & 2 & 3 \\
2 & 1 & 1 \\
2 & 1 & 2 \\
2 & 1 & 3 \\
2 & 2 & 1 \\
2 & 2 & 2 \\
2 & 2 & 3 
\end{array} } \right]
\end{align}

Each multivariate eigenfunction $\phi^{\ast}_j$ corresponds to the product of the univariate eigenfunctions whose indices corresponds to the elements of the $D$-tuple $\mathbb{S}_{j\cdotp}$, and each multivariate eigenvalue $\boldsymbol{\lambda}^{\ast}_j$ is a $D$-vector with elements that are the univariate eigenvalues whose indices correspond to the elements of the $D$-tuple $\mathbb{S}_{j\boldsymbol{\cdotp}}$. Thus, for $\mathbf{x}=\{x_d\}_{d=1}^D \in \Omega$ and $j=1,\ldots,m^{\ast}$, we have 
%
\begin{align}
\phi^{\ast}_j(\mathbf{x}) &= \prod_{d=1}^{D} \phi_{\mathbb{S}_{jd}}(x_d) = \prod_{d=1}^{D} \sqrt{\frac{1}{L_d}} \text{sin}\left(\sqrt{\lambda_{\mathbb{S}_{jd}}}(x_d+L_d)\right) \label{eigenvalue_multi} \\
%
\boldsymbol{\lambda}^{\ast}_j &= \left\{ \lambda_{\mathbb{S}_{jd}} \right\}_{d=1}^D =  \left\{ \left(\tfrac{\pi \mathbb{S}_{jd}}{2L_d}\right)^2 \right\}_{d=1}^D. \label{eigenfunction_multi}
\end{align}

\noindent The approximate covariance function is then represented as
%
\begin{equation}\label{approxcov_multi}
k(\mathbf{x},\mathbf{x}') \approx \sum_{j=1}^{m^{\ast}} 
S^{\ast}_{\theta}\left(\sqrt{\boldsymbol{\lambda}^{\ast}_j}\right)
\phi^{\ast}_j(\mathbf{x}) \phi^{\ast}_j(\mathbf{x}'),
\end{equation}

\noindent where $S^{\ast}_{\theta}$ is the spectral density of the $D$-dimensional covariance function (see Section 2.1). We can now write the approximate series expansion of the multivariate function $f$ as,
%
\begin{equation}\label{approxf_multi}
f(\mathbf{x}) \approx \sum_{j=1}^{m^{\ast}} 
\left( S^{\ast}_{\theta} \left(\sqrt{\boldsymbol{\lambda}^{\ast}_j} \right)\right)^{1/2}
\phi^{\ast}_j(\mathbf{x}) \beta_j, 
\end{equation}

@@

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equation (\ref{eigenvalue} and \ref{eigenfunction}) for the unidimensional case and in Equations (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{eigenvalue}), (\ref{eigenfunction}), (\ref{eigenvalue}) and (\ref{eigenfunction}), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Cov_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig1_legend.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different number of basis functions $m$. (right) Covariance functions of both regular GP model and HSGP model for different number of basis functions $m$.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Cov_L.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig2_legend.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different boundary factors $c$. (right) Covariance functions of both regular GP model and HSGP model for different boundary factors $c$.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.30, trim = 10mm 95mm 10mm 30mm, clip]{fig4_legend.pdf}}\\
\vspace{-1mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Post_part1.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Cov_part1.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Post_part2.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Cov_part2.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Post_part3.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Cov_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Post_part4.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Cov_part4.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Post_part5.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig4_Cov_part5.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 10mm 10mm 20mm, clip]{fig4_Post_part6.pdf}}
\subfigure{\includegraphics[scale=0.25, trim = 0mm 10mm 10mm 20mm, clip]{fig4_Cov_part6.pdf}}\\
\vspace{-3mm}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different boundary factors $c$. (right) Covariance functions of both regular GP model and HSGP model for different boundary factors $c$.}
  \label{fig4_Post_part1}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig5_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig5_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the regular GP model. (left) RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$. (right) RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$. }
  \label{fig5_MSE_vs_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_magnitud_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig6_legend.pdf}}
\caption{Estimated lengthscale (left) and marginal variance (right) parameters of both regular GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig6_lscale_vs_J}
\end{figure}

@@

\section{Univariate examples}

@@

This example consists of a simulated dataset with 250 ($i=1,\dots,n=250$) single draws of a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\mathbf{x}=\{x_i\}_{i=1}^{250}$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\mathbf{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.

The regular GP model for fitting this simulated dataset $\mathbf{y}$ can be written as follows,
%
\begin{eqnarray*}\label{eq:latentgp_simudata1}
\begin{split}
\mathbf{y} &= f(\mathbf{x}) + \boldsymbol{\epsilon} \\
\boldsymbol{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(\mathbf{x}) &\sim \mathcal{GP}(0, k(\mathbf{x}, \mathbf{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $I$ represents the identity matrix. The previous formulation corresponds to the latent form of a GP model, where $f(\mathbf{x})$ represents the underlying functional model to the noisy data, and it is modeled by a GP prior with a Mat{\'e}rn($\nu=3/2$) covariance function $k$. Saying that the function values $f(\mathbf{x})$ follows a GP model is equivalent to say that $f(\mathbf{x})$ are multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$.

A more computationally efficient formulation of a GP model with Gaussian likelihood, and for probabilistic inference using sampling methods such as HMC, would be its marginalized form,
%
\begin{equation*}\label{eq:marginalizedgp_simudata1}
\mathbf{y} \sim \mathcal{N}(0, K + \sigma^2 I )
\end{equation*}

@@

In the HSGP model, the latent function values $f(\mathbf{x})$ are approximated as (\ref{approxf}),
%
\begin{equation*}
f(\mathbf{x}) \approx \sum_{j}^m \left( S(\sqrt{\lambda_j})\right)^{1/2} \phi_j(\mathbf{x}) \beta_j,
\end{equation*}

\noindent with the spectral density $S$ (\ref{eq:specdens_32}) as a function of $\sqrt{\lambda_j}$,
%
\begin{eqnarray*}
S(\sqrt{\lambda_j})&=& 4\alpha^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \lambda_j)^{-2}, 
\end{eqnarray*}

\noindent and eigenvalues $\lambda_j$ (\ref{eigenvalue}) and eigenfunctions $\phi_j$ (\ref{eigenfunction})
%
\begin{eqnarray*}
\begin{split}
\lambda_j&=\left(\frac{j\pi}{2L}\right)^2, \\
\phi_j(\mathbf{x})&=\sqrt{\frac{1}{L}} \text{sin}\left(\sqrt{\lambda_j}(\mathbf{x}+L)\right).
\end{split}
\end{eqnarray*}
